Concentration Inequalities

### 2015-10-07 causal-bandit/observe_then_best.tex
works on bandit case for unknown q and standardizes some notation accross cases

It should also be possible to generalize to handle the case were the $q$'s are unknown, since we should be able to get a reasonable estimate for them while we are observing. They key will be how the resulting uncertainty in $m$ effects the bounds.

@@

\subsection{Generalizing to unknown q}

@@

\begin{itemize}
\item $N$ is number of variables.
\item $q_i$ is probability that $X_{i,t} = 1$ for any $t$
\item $s_i = \min\set{q_i, 1 - q_i}$
\item $h = T/2$
\item $\mu : \set{1,\ldots,N} \times \set{0,1} \to [0,1]$.
\item $I_t \in \set{1,\ldots,N} \times \set{0,1}$ is the intervention
\end{itemize}

Let $R_T$ be the simple regret after $T$ time steps, which is
\eq{
R_q(T) = \E\left[\mu^* - \mu_{I_T}\right]\,,
}
where $\mu^*$ is the mean payoff of the optimal intervention and $I_T$ is the intervention chosen in the $T$th round.
We propose a two part algorithm that is provably order optimal when its performance is measured in terms of the simple regret.

\begin{algorithm}[H]
\caption{Simple Regret Algorithm}\label{alg:simple}
\begin{algorithmic}[1]
\State {\bf Input:} $T, N$
\State $h = T/2$
\For{$t \in 1,\ldots,h$}
\State Do nothing and observe $X_{i,t}$ for $i \in \set{1,\ldots,N}$ and $r_t$
\EndFor
\State Compute for all $i \in \set{1,\ldots,N}$ and $j \in \set{0,1}$:
\eq{
\hat \mu_{i,j} = \frac{\sum_{t=1}^h \ind{X_{i,t} = j} r_t}{\sum_{t=1}^h \ind{X_{i,t} = j}}\,.
}
\State Compute $\hat q_i = \frac{1}{h} \sum_{t=1}^h X_{i,t}$
\State Compute $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$
\State Compute $\hat{s'} = sorted(\hat{s}) : \hat{s'}_1 \leq \hat{s'_2} \leq ... \leq \hat{s'}_N$
\State Compute $\hat m = \min\set{2 \leq i \leq N : \hat s'_{i} \geq 1/i}$
\State $i'(i) = $ the index of $\hat s_i$ in $\hat s'$
\State Compute $A$ as the subset of infrequently observed arms $\{(i,j):i'(i) \leq \hat m, j = \ind{\hat q_{i} \leq \frac{1}{2}} \}$ with $|A| = \hat m$

\For{$(i,j) \in A$}
\For{$t \in 1,\ldots,h/\hat m$}
\State Choose action $X_{i,t} = j$
\State Observe reward $r_t(X_{i,t}=j)$
\EndFor
\State Recompute $\hat \mu_{i,j} = \frac{\hat m}{h} \sum_{t=1}^{\hat m/h} r_t(X_{i,t}=j)$ 
\EndFor
\end{algorithmic}
\end{algorithm}

\begin{theorem}\label{thm:simple-regret}
Define $m = \min\set{2 \leq i \leq N : q_i \geq 1/i}$.
Then the algorithm given in \cref{alg:simple} satisfies
\eq{
r_q(T) \in O\left(\sqrt{\frac{m}{T} \log \left(\frac{NT}{m}\right)}\right)\,.
}
\end{theorem}

\begin{lemma}\label{lem:conc1}
$\displaystyle \P{\left|\hat q_i - q_i\right| \geq \sqrt{\frac{3q_i}{h} \log \frac{2}{\delta}}} \leq \delta$.
\end{lemma}

\begin{proof}
Let $Z_t = \ind{X_{i,t} = 1} \in \set{0,1}$.
Then
\eq{
\hat q_i = \frac{1}{h} \sum_{t=1}^{h} Z_t\,.
}
Now $Z_1,\ldots,Z_h$ is an i.i.d.\ sequence of Bernoulli random variables with with mean $q_i$. The result follows from the Chernoff bound.
\end{proof}

\begin{lemma}\label{lem:conc2}
Let $\delta >0$ and $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$ and
define $\hat m = \min\set{2 \leq i \leq N : \hat s_{(i)} \geq 1/i}$.
If
\eq{
h \geq 12m \log\frac{4N}{\delta}\,,
}
then
\eq{
\P{\hat m \leq 2m} \geq 1 - \delta\,.
}
\end{lemma}

\begin{proof}
It is easy to see that the worst case occurs when 
\eq{
q_i = \begin{cases}
0 & \text{if } i < m \\
\frac{1}{m} & \text{otherwise}\,.
\end{cases}
}
Now by \cref{lem:conc1} we have with probability at least $1 - \delta$ that 
\eq{
(\forall i) \qquad \left| \hat q_i - q_i\right| 
&\leq \sqrt{\frac{3q_i}{h} \log \frac{2N}{\delta}} \\
&= \sqrt{\frac{3}{mh} \log\frac{2N}{\delta}}\,. 
}
By assumption 
\eq{
h \geq 12m \log \frac{4N}{\delta}\,.
}
Therefore with probability at least $1 - \delta$ we have
\eq{
(\forall i) \qquad \left|\hat q_i - q_i\right| \leq \frac{1}{2m}\,.
}
Note that for $i$ with $q_i = 0$ it is easy to see that $\hat q_i = 0$ (the variance is zero). The above guarantee that
for $i$ with $q_i = 1/m$ we have $\hat q_i \in [1/(2m), 3/(2m)]$ for all $i$ with probability at least $1 - \delta$,
which implies that $\hat s_i \geq 1/(2m)$ for all $i$ where $q_i = 1/m$.
Therefore $\hat m \leq 2m$ with probability at least $1 - \delta$. 
\end{proof}

@@

for $(i,j)\in A$, the algorithm explicitly selects the action, $X_i = j$,  $\frac{h}{\hat m}$ times.
\eq {
\hat \mu_{i,j} = \frac{\hat m}{h} \sum_{t=1}^{\hat m/h} r_t(X_i = j)
}

@@

\eq {
\P{ \left|\hat \mu_{i,j} - \mu_{i,j}\right| > \epsilon } \leq 2\exp{-\frac{2h\epsilon^2}{\hat{m}}}
}

for $(i,j)\notin A$, the algorithm has observed the reward given $X_i = j$ at least $\frac{h}{\hat m}$ times.

@@

Let  $Z_{ij} = \sum_{t=1}^h \ind{X_i = j}$ and $t'_1 ... t'_{Z_{ij}} = {t:X_{i,t} = j}$

@@

\eq {
\P{\left| \hat \mu_{i,j} - \mu_{ij} \right| > \epsilon} = & \sum_{z=1}^\infty \P{Z_{ij}=z}\P{\left| \frac{1}{Z_{ij}} \sum_{t'=1}^{Z_{ij}}r_{t'} - \mu_{ij} \right| > \epsilon | Z_{ij} = z} \\
= & \sum_{z=1}^\infty \P{Z_{ij}=z}\P{\left| \frac{1}{z} \sum_{t'=1}^{z}r_{t'} - \mu_{ij} \right| > \epsilon } \\
\leq & \P{\left| \frac{\hat m}{h} \sum_{t'=1}^{h/\hat m}r_{t'} - \mu_{ij} \right| > \epsilon } \sum_{z=1}^\infty \P{Z_{ij}=z} \\
\leq & 2\exp{-\frac{2h\epsilon^2}{\hat{m}}}\\
}

Applying the union bound over all $2N$ actions.
\eq{
& \P{\exists (i,j): \left| \hat \mu_{i,j} - \mu_{i,j} \right| > \epsilon} \leq 4N\exp{-\frac{2h\epsilon^2}{\hat{m}}} \\
\implies & \P{\exists (i,j): \left| \hat \mu_{i,j} - \mu_{i,j} \right| > \sqrt{\frac{\hat m}{2h}\log{\frac{4N}{\delta}}}} \leq \delta \\
}

Now by \cref{lem:conc2}, 
\eq{
h \geq 12m \log \frac{4N}{\delta} \implies \P{\hat m > 2m} \leq \delta
}

Therefore if $h \geq 12m \log \frac{4N}{\delta}$, we have shown with probability at least $1 - 2\delta$ that

\eq{
(\forall i, j) \qquad \left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{m}{h} \log\frac{4N}{\delta}}\,.
}
Suppose $h < 12m \log \frac{4N}{\delta}$. Then
\eq{
\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq 1 \leq \sqrt{\frac{12m}{h} \log \frac{4N}{\delta}}\,.
}
Therefore
\eq{
\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{12m}{h} \log \frac{4N}{\delta}} \qquad \forall h\ \text{ with probability at least } 1-2\delta
}
Therefore 
\eq{
r_q(T) 
&\leq 2\delta + \sqrt{\frac{12m}{h} \log \frac{4N}{\delta}} \\
&= 2\delta + \sqrt{\frac{24m}{T} \log \frac{4N}{\delta}} \\
&\leq \frac{8m}{T} + \sqrt{\frac{24m}{T} \log \left(\frac{NT}{m}\right)}
}
as required.
\end{proof}

We now consider the case where $\boldsymbol{q}$ is not known in advance. Assume as before that we observe for $h/2$ timesteps. From the observations gained in this phase we estimate $\boldsymbol{q}$.

>>>>

Again comparing to the UCB bound of $\sqrt{KT}$, we expect the causal algorithm to do better if $K > m^{2/3}T^{1/3}$, or alternatively if $m \leq \frac{K^{3/2}}{\sqrt{T}}$

@@

\subsection{Generalizing to unknown q}
We now consider the case where $\boldsymbol{q}$, and thus $m$ is not known in advance. We begin by considering the simple regret

@@

\begin{itemize}
\item $N$ is number of variables.
\item $q_i$ is probability that $X_{i,t} = 1$ for any $t$
\item $s_i = \min\set{q_i, 1 - q_i}$
\item $\mu : \set{1,\ldots,N} \times \set{0,1} \to [0,1]$.
\item $I_t \in \set{1,\ldots,N} \times \set{0,1}$ is the intervention
\end{itemize}

Let $R_s(h)$ be the simple regret after $h$ time steps, which is
\eq{
R_s(h) = \E\left[\mu^* - \mu_{I_h}\right]\,,
}
where $\mu^*$ is the mean payoff of the optimal intervention and $I_h$ is the intervention chosen in the $h$th round.
We propose a two part algorithm that is provably order optimal when its performance is measured in terms of the simple regret.

\begin{algorithm}[H]
\caption{Simple Regret Algorithm}\label{alg:simple}
\begin{algorithmic}[1]
\State {\bf Input:} $h, N$
\For{$t \in 1,\ldots,h/2$}
\State Do nothing and observe $X_{i,t}$ for $i \in \set{1,\ldots,N}$ and $r_t$
\EndFor
\State Compute for all $i \in \set{1,\ldots,N}$ and $j \in \set{0,1}$:
\eq{
\hat \mu_{i,j} = \frac{\sum_{t=1}^{h/2} \ind{X_{i,t} = j} r_t}{\sum_{t=1}^{h/2} \ind{X_{i,t} = j}}\,.
}
\State Compute $\hat q_i = \frac{2}{h} \sum_{t=1}^{h/2} X_{i,t}$
\State Compute $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$
\State Compute $\hat{s'} = sorted(\hat{s}) : \hat{s'}_1 \leq \hat{s'_2} \leq ... \leq \hat{s'}_N$
\State Compute $\hat m = \min\set{2 \leq i \leq N : \hat s'_{i} \geq 1/i}$
\State $i'(i) = $ the index of $\hat s_i$ in $\hat s'$
\State Compute $A$ as the subset of infrequently observed arms $\{(i,j):i'(i) \leq \hat m, j = \ind{\hat q_{i} \leq \frac{1}{2}} \}$ with $|A| = \hat m$

\For{$(i,j) \in A$}
\For{$t \in 1,\ldots,h/2\hat m$}
\State Choose action $X_{i,t} = j$
\State Observe reward $r_t(X_{i,t}=j)$
\EndFor
\State Recompute $\hat \mu_{i,j} = \frac{2\hat m}{h} \sum_{t=1}^{ h/2\hat m} r_t(X_{i,t}=j)$ 
\EndFor
\end{algorithmic}
\end{algorithm}

\begin{theorem}\label{thm:simple-regret}
Define $m = \min\set{2 \leq i \leq N : q_i \geq 1/i}$.
Then the algorithm given in \cref{alg:simple} satisfies
\eq{
R_s(h) \in O\left(\sqrt{\frac{m}{h} \log \left(\frac{NT}{m}\right)}\right)\,.
}
\end{theorem}

\begin{lemma}\label{lem:conc1}
$\displaystyle \P{\left|\hat q_i - q_i\right| \geq \sqrt{\frac{6q_i}{h} \log \frac{2}{\delta}}} \leq \delta$.
\end{lemma}

\begin{proof}
Let $Z_t = \ind{X_{i,t} = 1} \in \set{0,1}$.
Then
\eq{
\hat q_i = \frac{2}{h} \sum_{t=1}^{h/2} Z_t\,.
}
Now $Z_1,\ldots,Z_{h/2}$ is an i.i.d.\ sequence of Bernoulli random variables with with mean $q_i$. The result follows from the Chernoff bound.
\end{proof}

\begin{lemma}\label{lem:conc2}
Let $\delta >0$ and $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$ and
define $\hat m = \min\set{2 \leq i \leq N : \hat s_{(i)} \geq 1/i}$.
If
\eq{
h \geq 24m \log\frac{4N}{\delta}\,,
}
then
\eq{
\P{\hat m \leq 2m} \geq 1 - \delta\,.
}
\end{lemma}

\begin{proof}
It is easy to see that the worst case occurs when 
\eq{
q_i = \begin{cases}
0 & \text{if } i < m \\
\frac{1}{m} & \text{otherwise}\,.
\end{cases}
}
Now by \cref{lem:conc1} we have with probability at least $1 - \delta$ that 
\eq{
(\forall i) \qquad \left| \hat q_i - q_i\right| 
&\leq \sqrt{\frac{6q_i}{h} \log \frac{2N}{\delta}} \\
&= \sqrt{\frac{6}{mh} \log\frac{2N}{\delta}}\,. 
}
By assumption 
\eq{
h \geq 24m \log \frac{4N}{\delta}\,.
}
Therefore with probability at least $1 - \delta$ we have
\eq{
(\forall i) \qquad \left|\hat q_i - q_i\right| \leq \frac{1}{2m}\,.
}
Note that for $i$ with $q_i = 0$ it is easy to see that $\hat q_i = 0$ (the variance is zero). The above guarantee that
for $i$ with $q_i = 1/m$ we have $\hat q_i \in [1/(2m), 3/(2m)]$ for all $i$ with probability at least $1 - \delta$,
which implies that $\hat s_i \geq 1/(2m)$ for all $i$ where $q_i = 1/m$.
Therefore $\hat m \leq 2m$ with probability at least $1 - \delta$. 
\end{proof}

@@

for $(i,j)\in A$, the algorithm explicitly selects the action, $X_i = j$,  $\frac{h}{2\hat m}$ times.
\eq {
\hat \mu_{i,j} = \frac{2\hat m}{h} \sum_{t=1}^{h/2\hat m} r_t(X_i = j)
}

@@

\eq {
\P{ \left|\hat \mu_{i,j} - \mu_{i,j}\right| > \epsilon } \leq 2\exp{-\frac{h\epsilon^2}{\hat{m}}}
}

for $(i,j)\notin A$, the algorithm has observed the reward given $X_i = j$ at least $\frac{h}{2\hat m}$ times.

@@

Let  $Z_{ij} = \sum_{t=1}^{h/2} \ind{X_i = j}$ and $t'_1 ... t'_{Z_{ij}} = {t:X_{i,t} = j}$

@@

\eq {
\P{\left| \hat \mu_{i,j} - \mu_{ij} \right| > \epsilon} = & \sum_{z=1}^\infty \P{Z_{ij}=z}\P{\left| \frac{1}{Z_{ij}} \sum_{t'=1}^{Z_{ij}}r_{t'} - \mu_{ij} \right| > \epsilon | Z_{ij} = z} \\
= & \sum_{z=1}^\infty \P{Z_{ij}=z}\P{\left| \frac{1}{z} \sum_{t'=1}^{z}r_{t'} - \mu_{ij} \right| > \epsilon } \\
\leq & \P{\left| \frac{2\hat m}{h} \sum_{t'=1}^{h/2\hat m}r_{t'} - \mu_{ij} \right| > \epsilon } \sum_{z=1}^\infty \P{Z_{ij}=z} \\
\leq & 2\exp{-\frac{h\epsilon^2}{\hat{m}}}\\
}

Applying the union bound over all $2N$ actions.
\eq{
& \P{\exists (i,j): \left| \hat \mu_{i,j} - \mu_{i,j} \right| > \epsilon} \leq 4N\exp{-\frac{h\epsilon^2}{\hat{m}}} \\
\implies & \P{\exists (i,j): \left| \hat \mu_{i,j} - \mu_{i,j} \right| > \sqrt{\frac{\hat m}{h}\log{\frac{4N}{\delta}}}} \leq \delta \\
}

Now by \cref{lem:conc2}, 
\eq{
h \geq 24m \log \frac{4N}{\delta} \implies \P{\hat m > 2m} \leq \delta
}

Therefore if $h \geq 24m \log \frac{4N}{\delta}$, we have with probability at least $1 - 2\delta$ that

\eqn{
\label{eqn:joint_bound_estimates_m}
(\forall i, j) \qquad \left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{2m}{h} \log\frac{4N}{\delta}}\, \text { and } \hat{m} \leq 2m
}
Suppose $h < 24m \log \frac{4N}{\delta}$. Then
\eq{
\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq 1 \leq \sqrt{\frac{2m}{h} \log \frac{4N}{\delta}}\,.
}
Therefore
\eq{
\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{24m}{h} \log \frac{4N}{\delta}} \qquad \forall h\ \text{ with probability at least } 1-2\delta
}
Therefore 
\eq{
R_s(h) 
&\leq 2\delta + \sqrt{\frac{24m}{h} \log \frac{4N}{\delta}} \\
&\leq \frac{8m}{h} + \sqrt{\frac{24m}{h} \log \left(\frac{Nh}{m}\right)}
}
as required.
\end{proof}

\subsubsection{Bounding the Bandit Regret}

\begin{itemize}
\item Use $24Nlog(4N/\delta)$ samples to estimate $m$
\item If $\hat m > \frac{N^{3/2}}{\sqrt{T}}$ then stop and do UCB with remaining rounds.
\item Else use causal explore-exploit and let the exploration time $h = T^{2/3}\hat{m}^{1/3}log(TK)^{1/3}$ 
\end{itemize}

Since we don't know $\boldsymbol{q}$, and thus $m$, we can't set the exploration time $h$ in advance based on $m$ as we did for the known $\boldsymbol{q}$ case. Instead we first use $24Nlog(4N/\delta)$ samples to estimate $m$. If $\hat m < \frac{N^{3/2}}{\sqrt{T}}$ we will continue with the causal algorithm and let the total exploration time $h = T^{2/3}\hat{m}^{1/3}log(TK)^{1/3}$.

Since $m \leq N$, by \cref{eqn:joint_bound_estimates_m}, we have that with probability at least $1-2\delta$,

\eqn{
\label{eqn:bounds_hold}
(\forall i, j) \qquad \left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{\hat m}{h} \log\frac{4N}{\delta}}\, \text { and } \hat{m} \leq 2m
}

So the regret in this case is bounded by,

\eq {
R(T) & =  2\delta E\left[R(T)|\text{ not } \cref{eqn:bounds_hold}\right] + (1-2\delta)E[R(T)| \cref{eqn:bounds_hold}]  \\
& \leq 2\delta T + \left(h + T\sqrt{\frac{\hat m}{h} \log\frac{4N}{\delta}} \right)\\
 &  \leq \frac{4 \hat m^{1/3}T}{h}  + h + T \left( \sqrt{\frac{\hat m}{h} \log \left(\frac{2Nh}{\hat m^{1/3}}\right)}\right), \text { letting } \delta = \frac{2 \hat m^{1/3}}{h} \\
&  \leq 4\left(\frac{T}{log(TK)}\right)^{1/3} + h + T \left( \sqrt{\frac{\hat m}{h} \log \left(TK\right)}\right) \\
&  \leq 4\left(\frac{T}{log(TK)}\right)^{1/3} + 2T^{2/3}(2m)^{1/3}log(TK)^{1/3}  \\
&  \leq 8T^{2/3}m^{1/3}log(TK)^{1/3}  \\
}

Now make sure I have all the cases right - then have a go with an implementation.

\subsubsection{ Older Attempts}
We now consider the case where $\boldsymbol{q}$ is not known in advance. Assume as before that we observe for $h/2$ timesteps. From the observations gained in this phase we estimate $\boldsymbol{q}$.
